---
title: "Visualizing Deadly Syrian Migrant Routes"
author: "John"
date: "August 11, 2017"
output: html_document
---

### Deadly Syrian Migration Routes to Europe 

This notebook explores the migration data set and examines the most deadly routes for Syrians going to Europe.  

We'll start by exploring the variables in the data set, doing a little data carpentry, and look for key insights.   Also if anyone else has any ideas on how to improve the analysis please leave your comments below.  Thanks in advance! 

Let's get started. 

```{r}
suppressPackageStartupMessages({library(ggplot2) 
library(readr) 
library(tidyverse)
library(dplyr)
library(gridExtra)
library(grid)
library(ggthemes)
library(RColorBrewer)
library(ggfortify)
library(rworldmap)
library(leaflet)})

```

###**Part 1: Data Cleaning and Initial Exploration** 

Let's clean up some of the NAs are located in the dataframe 

```{r}
missing <- read_csv("C:/Users/John/projects/Visualizations/MissingMigrantsProject.csv")
head(missing)
str(missing)
summary(missing) 

# I'm going to simplify the column names just a bit 
colnames(missing) <- c('id', 'cause_death', 'origin', 'nationality', 'missing', 'dead', 'incident_location', 'date', 'source', 'reliability', 'lat', 'lon')

# check out your new column names
names(missing)
```

```{r}
# we have a lot of NAs for columns with missing and dead counts and most other columns look pretty good.  
# we will clean this up a little bit focusing first on columns with numbers.  

missing[c("missing", "dead", "date", "lat", "lon")][is.na(missing[c("missing", "dead", "date", "lat", "lon")])] <- 0

summary(missing)
str(missing)
```

Let's first clean up the date column.  

```{r}
library(lubridate) 

#output looks like  we are working with a day / month / year format
missing$date <- dmy(missing$date)
str(missing$date) # view changes 

summary(missing) # looks like there's a few dates with NAs that were introduced do to some messy data. 
# Let's clean up our other columns and we'll come back to this.  
# Dropping these 9 rows might be easiest if there's something odd going on with these dates. 

```
Now examine the NA situation in our data set.  We'll summarize the NAs in each column.  Nationality seems to be the column containing the most NAs.  We can view this in a number of different ways.  

```{r}
library(dplyr)
missing %>% 
    group_by(origin) %>%   
    summarise_each(funs(sum(is.na(.))))


# That is a large proportion of the number of records in the data
sum(is.na(missing$origin))
sum(is.na(missing$source))
sum(is.na(missing$reliability))
sum(is.na(missing$nationality))
```

For our analysis I'd rather have unknown as a value for some of these columns with NAs or missing values.  For some of these we're dealing with locations and nationalities.  We can assume we just don't know where these people are from or where they went missing/died.  


```{r}
# This code below just cleans up our NA situation by replacing the values, ultimately putting "unknown" in columns with characters that had many NAs
missing[c("origin")][is.na(missing[c("origin")])] <- 0
missing[c("nationality")][is.na(missing[c("nationality")])] <- 0
missing[c("incident_location")][is.na(missing[c("incident_location")])] <- 0
missing[c("reliability")][is.na(missing[c("reliability")])] <- 0
missing[c("source")][is.na(missing[c("source")])] <- 0

missing$origin <- gsub("0", "Unknown", missing$origin)
missing$nationality <- gsub("0", "Unknown", missing$nationality)
missing$incident_location <- gsub("0", "Unknown", missing$incident_location)
missing$reliability <- gsub("0", "Unknown", missing$reliability)
missing$source <- gsub("0", "Unknown", missing$source)

# for simplicity I'm just going to drop those 9 NAs in the date column. 
missing <- missing %>% filter(!is.na(date))
```

Now let's examine the regions in the data set where people go missing or die.  The code below shows that most people die or go missing around the Mediterranean. This seems to match what we're hearing in the news and helps contextualize all the stories we hear about refugees drowning in the Mediterranean sea.   

```{r}
regions <- missing %>% 
    group_by(incident_location) %>% 
    summarise(sum(missing), sum(dead))

regions <- as.data.frame(regions)
colnames(regions) <- c('incident_location', 'missing', 'dead')
regions <- arrange(regions, dead)
regions$incident_location <- factor(regions$incident_location, levels = regions$incident_location[order(regions$dead)])

regions
```

Let's go ahead and plot this table.  

```{r}
library(ggplot2)

# Look at the distribution of deaths by region 
p <- ggplot(regions, aes(x=incident_location, y=dead)) + 
  geom_bar(stat ='identity') + 
  coord_flip() + 
  ggtitle("migrant deaths recorded by region") +
  theme_fivethirtyeight() 

p
```

Now we plot the missing persons (migrants) by region where they went missing.  The Mediterranean region stands out. 

The plots clearly show the Mediterranean as a location of interest. The incidents of missing persons and deaths in the Mediterranean are much greater than any other location throughout the world.  North Africa is another region that sticks out here with a large number of recorded deaths.  
```{r}
# clean and plot missing persons by region 
missing_sums <- subset(regions, missing >= 1)

missing_sums  <- arrange(missing_sums , missing)
missing_sums$incident_location <- factor(missing_sums$incident_location, levels = missing_sums$incident_location[order(missing_sums$missing)])

p1 <- ggplot(missing_sums, aes(x=incident_location, y=missing)) + 
  geom_bar(stat ='identity') + 
  coord_flip() + 
  ggtitle("missing persons recorded by region") +
  theme_fivethirtyeight() 

p1
```

After further exploration of the original csv file we notice  a large number of records for North Africa were left blank.  That's why they were filled with "unknown" earlier in this workbook.  Going forward we are going to assume that these "unknown" values which were left blank in the original data set should be "unverified".  We're going to swap out "unknown" for "unverified".


```{r}
# order plot
positions <- c("Mediterranean", "North Africa", "U.S./Mexico Border", "Southeast Asia", 
               "Horn of Africa", "Central America incl. Mexico", "Sub-Saharan Africa", 
               "Europe", "Caribbean", "South America", "North America", "East Asia", "Unknown")

p2 <- ggplot(missing, aes(x=factor(incident_location), y=dead, fill=factor(reliability))) +
     geom_bar(stat = 'identity') +
     coord_flip() +
     ggtitle("deaths recorded by region") +
     scale_x_discrete(limits = positions)
     
p2    

# we have an "unknown" value in this column.  For this analysis "unverified" is more appropriate.  
missing$reliability <- gsub("Unknown", "Unverified", missing$reliability)


# We also notice that we have a spelling issue with "partially verified" when 
# the data set was created. One was created with caps and one without. We'll want to fix this.  

missing$reliability <- gsub("Partially verified", "Partially Verified", missing$reliability)


#plot again to view the updates.  This is looking better. 
p3 <- ggplot(missing, aes(x=factor(incident_location), y=dead, fill=factor(reliability))) +
     geom_bar(stat = 'identity') +
     coord_flip() +
     ggtitle("deaths recorded by region") +
     scale_x_discrete(limits = positions)
     
p3    
```


```{r}

# plot missing persons and deaths over time 

#deaths by date
p4 <- ggplot(missing, aes(x=date, y=dead)) + 
  geom_line() +
  ggtitle("deaths by date")
p4

# missing persons by date
p5 <- ggplot(missing, aes(x=date, y=missing)) + 
  geom_line() +
  ggtitle("missing persons by date")
p5

#Let's plot the dead variable over time and show the reliablity feature.  
# We can see that the unverified or unknown data was documented in 2016.
p6 <- ggplot(missing, aes(x=date, y=dead)) + 
    geom_point(aes(col=reliability), alpha=0.6) +
    ggtitle('reliability of recorded deaths')
p6


```

View number or records for region of origin. Origin seems to be a good variable in the data set.  Not only do we see this here when we count the records, but we also notice that the IOM focuses on this variable when presenting their data to the public. That's a good sign. 

```{r}
origin_reg <- missing %>% group_by(origin) %>% count()

origin_reg <- as.data.frame(origin_reg)
head(origin_reg, 20)
```

The Nationality column is a pretty messy column.  I'm liking the region of origin variable after exploring how Nationality is tracked in the data set.  

We've got about 220 categories of nationalities.  Some are duplicates like 'Mexico' and 'Mexican'.  Others have some ethnic suggestions to them such as 'Myanmar' and 'Myanmar (Rohingya)'.  Others vary like categories for Syria such as Syria', 'Syrian' and 'Syria Arab Republic'... even 'African' and 'Sub-Saharan'...wondering how accurate these categories actually are...could a sub-saharan nationality been classified as 'African' when the data set was created? I'm even seeing some columns with multiple nationalities and ethnicities listed. Cleaning this up is going to take some time. There are also a lot of unknowns in this column - 1567 to be exact.  Still hard to tell how much value can be derived from this column given how messy it is and the amount of unknowns. 
```{r}
nations <- missing %>% 
    group_by(nationality) %>%   
    count()
nations <- as.data.frame(nations)
head(nations, 20) 
```

### Let's focus on the tricky cause of death column. Here we see hundreds of causes of death. 

The cause of death column has 290 different categorical variables.  There appear to be some variables that are the same with different spellings.  Others appear to have a lot of detail, but this detail is going to be difficult for anyone to make sense of when looking for patterns in the data.  It would be great if they used some type of standard classification method here to make sense of what's going on.  Going forward we're going to need to clean this up, address the spelling issues, and collapse certain categories.  

Before we get started I found this helpful resource from IOM.  
resource: https://missingmigrants.iom.int/sites/default/files/gmdac_data_briefing_series_issue4.pdf 

On page 5 of their report, the missing migrants project seems to have collapsed or at least they are presenting the causes of death in a limited number of categories.  This is a helpful start.  When collapsing variable values we'll try use this structure as a starting point and see how far we get.  

Before we start cleaning let's look at all the cause of death values.  
```{r}
count <- missing %>% 
    group_by(cause_death) %>%  
    count()

count <- as.data.frame(count)
head(count, 20) # limit the print because there are too many 
```
As a first step let's just take out the records we're not interested.  These include records from other regions.  Cleaning up that cause of death column is going to take some time and since we're focused on the Mediterranean region, let's first subset the data.  

The incident location will be a key variable allowing us to examine the Mediterranean region in more detail.  
```{r}
# subset the dataframe to the following regions: Europe, Mediterranean, Middle East, North Africa
missing_med <- subset(missing, incident_location == "Europe" | incident_location == "Mediterranean" | incident_location == "Middle East" | incident_location =="North Africa")

#check the data set.  We're down to 1415 variables.  
str(missing_med)
regions <- missing_med %>% group_by(incident_location) %>% summarise(sum(missing), sum(dead))
regions
```
### Below are the new cause of death categories we'll use going forward: 
These are pretty similar to what's in the data set. 

 1. Drowning
 2. Medical 
 3. Asphyxiation 
 4. Vehicle Accident
 5. Train Accident
 6. Violence, Assault, Murder
 7. Exhaustion Starvation, Dehydration, Exposure
 8. Sexual Assault
 9. Accident
 10. Unknown


```{r}
# this was a bit tedious to reclassify these as appropriate. 

cleanup <- function(df) {
    df <- gsub("Boat fire","Accident",df)
    df <- gsub("Asphyxiation and crushing", "Asphyxiation",df)
    df <- gsub("Beat-up and killed", "Violence, Assault, Murder",df)
    df <- gsub("Burned to death hiding in truck", "Vehicle Accident",df)
    df <- gsub("Burns and Suffocation", "Asphyxiation",df)
    df <- gsub("Burns from cooking gas explosion in connection house in Libya", "Accident",df)
    df <- gsub("Clubbed/beaten to death","Violence, Assault, Murder",df)
    df <- gsub("Crushed","Accident",df)
    df <- gsub("Crushed / drowning","Drowning",df)
        df <- gsub("Boat","Accident",df)
    df <- gsub("Accident / drowning","Accident",df)
    df <- gsub("Accident by bus on ferry","Accident",df)
    df <- gsub("Accident by pallets","Accident",df)
    df <- gsub("Accident to death","Accident",df)
    df <- gsub("Dehydration","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Dehydration and exposure to the elements","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Dehydration Harsh_weather_lack_of_adequate_shelter","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Dehydration Harsh_weather_lack_of_adequate_shelter Suffocation Excessive_physical_abuse Sexual_abuse","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Dehydration Suffocation Vehicle_Accident","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Dehydration Vehicle_Accident Excessive_physical_abuse","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Dehydration, Starvation","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Died of unknown cause in hospital shortly after rescue","Unknown",df)
    df <- gsub("drowning","Drowning",df)
    df <- gsub("Drowning after being thrown overboard by other passengers","Drowning",df)
    df <- gsub("Drowning or suffocation in hull","Drowning",df)
    df <- gsub("Drowning, Asphyxiation","Drowning",df)
    df <- gsub("Drowning, Other","Drowning",df)
    df <- gsub("Drowning, Trampling","Drowning",df)
    df <- gsub("Drowning. Boat collided with ferry","Drowning",df)
    df <- gsub("Electrocuted on train","Train Accident",df)
    df <- gsub("Electrocution","Accident",df)
    df <- gsub("Electrocution on railway","Train Accident",df)
    df <- gsub("Excessive_physical_abuse","Violence, Assault, Murder",df)
    df <- gsub("Excessive_physical_abuse Sexual_abuse","Sexual Assault",df)
    df <- gsub("Excessive_physical_abuse Shot_or_Stabbed","Violence, Assault, Murder",df)
    df <- gsub("Exposure","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Exposure, Hyperthermia","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Exposure. Died upon entry to refugee camp.","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Fell from boat","Accident",df)
    df <- gsub("Fell from train","Train Accident",df)
    df <- gsub("Fell from truck","Vehicle Accident",df)
    df <- gsub("Fuel burns","Accident",df)
    df <- gsub("Fuel Inhalation","Asphyxiation",df)
    df <- gsub("Gylcemic crisis (Diabetic, medicine thrown overboard)","Medical",df)
    df <- gsub("Harsh conditions","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Harsh_weather_lack_of_adequate_shelter","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Harsh_weather_lack_of_adequate_shelter Excessive_physical_abuse","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Harsh_weather_lack_of_adequate_shelter Excessive_physical_abuse Sexual_abuse","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Harsh_weather_lack_of_adequate_shelter Other","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Harsh_weather_lack_of_adequate_shelter Suffocation Vehicle_Accident","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Harsh_weather_lack_of_adequate_shelter Vehicle_Accident","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Head injury","Accident",df)
    df <- gsub("Head injury from fall","Accident",df)
    df <- gsub("Head trauma (hit by boat propeller)","Accident",df)
    df <- gsub("Hi by truck","Vehicle Accident",df)
    df <- gsub("Hit by car","Vehicle Accident",df)
    df <- gsub("Hit by train","Train Accident",df)
    df <- gsub("Hit by truck","Vehicle Accident",df)
    df <- gsub("Hit by vehicle","Vehicle Accident",df)
    df <- gsub("Hit by Vehicle","Vehicle Accident",df)
    df <- gsub("Homicide, likely by asphyxiation","Violence, Assault, Murder",df)
    df <- gsub("Hunger, fatigue","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Hyperthermia","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Hyperthermia, Abandoned by smugglers in the desert","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Hyperthermia, starvation","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Hypothermia, Exhaustion","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Hypothermia, Malnutrition","Accident",df)
    df <- gsub("Inhalation of toxic fumes from boat engine","Accident",df)
    df <- gsub("Injured from a fight","Violence, Assault, Murder",df)
    df <- gsub("Killed","Violence, Assault, Murder",df)
    df <- gsub("NA","Unknown",df)
    df <- gsub("Presumed Drowning","Drowning",df)
    df <- gsub("Hypothermia","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Unknown (found on motorway)","Unknown",df)
    df <- gsub("Unspecified location between North Africa and Italy. Body brought to Calabria.","Unknown",df)
    df <- gsub("Starvation, Exhaustion, Starvation, Dehydration, Exhaustion, Starvation, Dehydration, Exposure, Exhaustion","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Exhaustion, Starvation, Dehydration, Exposure, Abandoned by smugglers in the desert","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Toxic fumes/asphyxiation","Asphyxiation",df)
    df <- gsub("Shot","Violence, Assault, Murder ",df)
    df <- gsub("Plane stowaway","Unknown",df)
    df <- gsub("Suffocation, Trampled","Asphyxiation",df)
    df <- gsub("Pulmonary edema","Medical",df)
    df <- gsub("Suffocation","Asphyxiation",df)
    df <- gsub("Unknown (body recovered from boat)","Unknown",df)
    df <- gsub("Unknown (found dead on top of train)","Unknown",df)
    df <- gsub("Unknown (body recovered from boat)","Unknown",df)
    df <- gsub("Presumed asphyxiation","Asphyxiation",df)
    df <- gsub("Vehicle accident","Vehicle Accident",df)
    df <- gsub("Gylcemic crisis (Diabetic, medicine thrown overboard)","Medical",df)
    df <- gsub("Unknown, plane stowaway","Unknown",df)
    df <- gsub("Violent robbery","Violence, Assault, Murder",df)
    df <- gsub("Meningitis","Medical",df)
    df <- gsub("Exhaustion, Starvation, Dehydration, Exposure, hypothermia","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Accident from fall","Accident",df)
    df <- gsub("Accident on railway","Train Accident",df)
    df <- gsub("Asphyxiation Vehicle_Accident","Asphyxiation",df)
    df <- gsub("Asphyxiation, Other","Asphyxiation",df)
    df <- gsub("Drowning. Accident collided with ferry","Drowning",df)
    df <- gsub("Drowning. Accident collided with ferry","Unknown",df)
    df <- gsub("Exhaustion, Exhaustion, Starvation, Dehydration, Exposure, Exhaustion, Starvation, Dehydration, ","Exhaustion, Starvation, Dehydration, Exposure",df)
    df <- gsub("Likely Drowning ","Drowning",df)
    df <- gsub("Lung infection","Medical",df)
    
    return(df)
}

# this function helps, but there's still a lot of categories that need cleaning.  
missing_med$cause_death <- cleanup(missing_med$cause_death)


# further cleaning up this column using using a more flexible technique. 
missing_med$cause_death[startsWith(missing_med$cause_death, "Exhaustion")] <- "Exhaustion, Starvation, Dehydration, Exposure"
missing_med$cause_death[startsWith(missing_med$cause_death, "Starvation")] <- "Exhaustion, Starvation, Dehydration, Exposure"
missing_med$cause_death[startsWith(missing_med$cause_death, "Violence")] <- "Violence, Assault, Murder"
missing_med$cause_death[startsWith(missing_med$cause_death, "Sickness")] <- "Medical"
missing_med$cause_death[startsWith(missing_med$cause_death, "Gylcemic")] <- "Medical"
missing_med$cause_death[startsWith(missing_med$cause_death, "Head trauma")] <- "Medical"
missing_med$cause_death[startsWith(missing_med$cause_death, "Likely suffocation")] <- "Asphyxiation"
missing_med$cause_death[startsWith(missing_med$cause_death, "Murdered")] <- "Violence, Assault, Murder"
missing_med$cause_death[startsWith(missing_med$cause_death, "Likely Drowning")] <- "Drowning"
missing_med$cause_death[startsWith(missing_med$cause_death, "On board violence")] <- "Unknown"
missing_med$cause_death[startsWith(missing_med$cause_death, "Plane Stowaway")] <- "Unknown"
missing_med$cause_death[startsWith(missing_med$cause_death, "Presumed dehydration")] <- "Exhaustion, Starvation, Dehydration, Exposure"
missing_med$cause_death[startsWith(missing_med$cause_death, "Other")] <- "Unknown"
missing_med$cause_death[startsWith(missing_med$cause_death, "Presumed violence")] <- "Violence, Assault, Murder"
missing_med$cause_death[startsWith(missing_med$cause_death, "Probable Drowning")] <- "Drowning"
missing_med$cause_death[startsWith(missing_med$cause_death, "Respiratory problem")] <- "Medical"
missing_med$cause_death[startsWith(missing_med$cause_death, "Road accident")] <- "Vehicle Accident"
missing_med$cause_death[startsWith(missing_med$cause_death, "Sexual_abuse")] <- "Sexual Assault"
missing_med$cause_death[startsWith(missing_med$cause_death, "Tekeze River, near Himora, Ethiopia")] <- "Unknown"
missing_med$cause_death[startsWith(missing_med$cause_death, "Train accident")] <- "Sexual Assault"
missing_med$cause_death[startsWith(missing_med$cause_death, "Stabbed")] <- "Violence, Assault, Murder"
missing_med$cause_death[startsWith(missing_med$cause_death, "Truck crash")] <- "Vehicle Accident"
missing_med$cause_death[startsWith(missing_med$cause_death, "Unclear")] <- "Unknown"
missing_med$cause_death[startsWith(missing_med$cause_death, "Vehicle incident")] <- "Vehicle Accident"
missing_med$cause_death[startsWith(missing_med$cause_death, "Vehicle_Accident")] <- "Vehicle Accident"
missing_med$cause_death[startsWith(missing_med$cause_death, "Undernourished")] <- "Exhaustion, Starvation, Dehydration, Exposure"
missing_med$cause_death[startsWith(missing_med$cause_death, "Unknown")] <- "Unknown"
missing_med$cause_death[startsWith(missing_med$cause_death, "unknown")] <- "Unknown"

# We've got some leftover NAs encoded as <NA> in the data.  The normal na removal code isn't 
# working so this bit tranforms things and changes these to unknown.  
missing_med$cause_death[is.na(missing_med$cause_death)] <- 0
missing_med$cause_death <- gsub(0,"Unknown", missing_med$cause_death)


# View new cause of death categories and sum the missing persons and death counts 
missing_med %>% 
    group_by(cause_death) %>% 
    summarise(sum(missing), sum(dead))

```

### Part 2: Data Exploration 

In this second part we'll focus more on exploring our data and less on cleaning and data manipulation.  We will start by looking at the data spatially.  The data set contains latitude and longitude coordinates so we'll start to visualize and map the data to see if we find any useful insights.  
```{r}
# plot the lat and lon cords and show regional classifications 
location_plot <- ggplot(missing_med, aes(x = lat, y = lon), col=incident_location) + geom_point(aes(col=incident_location))
location_plot
```


```{r}
# look at our missing persons records on a map 
med_missing_p <- subset(missing_med, missing >= 1)

# specify color palette for map points
pal <- colorNumeric(palette = colorRamp(c("#a669f4", "#0c000f"), interpolate = "spline"),
                   reverse = FALSE,  
                   domain = NULL)

leaflet(med_missing_p) %>% addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat, weight = 4, color = ~pal(missing),
    radius = ~missing * 500, popup = ~incident_location) %>% 
      addLegend("bottomright", colors= "#a669f4", labels="Missing'", title="Missing Migrants <br>in the Greater <br>Mediterranean <br>Region %")
```

Now let's look at deaths around the Mediterranean Reion 
```{r}
# view the location of deaths 
med_dead <- subset(missing_med, dead >= 1)

# specify color palette for map points
 

leaflet(med_dead) %>% addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat, weight = 4, color = ~"#ffa500", stroke = TRUE,
    radius = ~dead * 500, popup = ~incident_location) %>% 
      addLegend("bottomright", colors= "#ffa500", labels="Deaths'", title="Migrant Deaths <br>in the Greater <br>Mediterranean <br>Region %")
```

Missing and Dead Migrants in the Greater Mediterranean Region 
```{r}


leaflet() %>% 
  addTiles() %>%
  addCircles(data = med_dead, group = "Deaths", lng = ~lon, lat = ~lat, weight = 4, color = ~"#ffa500", stroke = TRUE,
    radius = ~dead * 500, popup = ~incident_location) %>%
  addCircles(data = med_missing_p, group = "Missing", lng = ~lon, lat = ~lat, weight = 4, color = ~"#a669f4", stroke = TRUE,
    radius = ~missing * 500, popup = ~incident_location) %>% 
    # Layers control
  addLayersControl(
    baseGroups = c("Deaths (default)", "Missing"),
    options = layersControlOptions(collapsed = FALSE)) %>%
      addLegend("bottomright", colors= "#a669f4", values = ~missing, labels="missing", title="Missing Migrants") %>%
      addLegend("bottomright", colors= "#ffa500", values = ~dead, labels="dead", title="Migrant Deaths")




```
### Part 3: Adding New Features to Analyze Mediterranean Migrant Routes 

Now we are going to add another feature called migration route. This new feature is closely related to incident location. However, we're breaking thigs down a little more based on information from Frontex (Europe's border patrol agency) and IOM defined migration routes. According to Frontex and IOM several migration routes exist around the Mediterranean Sea.  Here's some helpful links for more background on the routes: 
http://frontex.europa.eu/trends-and-routes/migratory-routes-map/ 
http://migration.iom.int/europe/ 

I think these breaks represent a good start, but they could probably be improved if you had more specialized knowledge of the routes.
```{r}
# define the space for routes using the lat and lon coords and incident location 
missing_med$route <- ifelse((missing_med$lat <= 40) & (missing_med$lat >= 30) & (missing_med$lon <= 5) & (missing_med$lon >= -20), "Western Mediterranean Route",
                      ifelse((missing_med$lat <= 41) & (missing_med$lat >= 28) & (missing_med$lon >= 5) & (missing_med$lon <= 22), "Central Mediterranean Route", 
                      ifelse((missing_med$lat <= 35) & (missing_med$lat >= 30) & (missing_med$lon >= 20 ) & (missing_med$lon <= 35), "Apulia and Calabria Route",
                      ifelse((missing_med$lat <= 41) & (missing_med$lat >= 35) & (missing_med$lon >= 20) & (missing_med$lon <= 80), "Eastern Mediterranean Route", 
                      ifelse((missing_med$lat >= 41) & (missing_med$lat <= 47) & (missing_med$lon >= 18) & (missing_med$lon <= 28), "Western Balkan Route", 
                      ifelse((missing_med$incident_location == "North Africa") | (missing_med$incident_location == "Sub-Saharan Africa"), "Africa Route", 
                      ifelse((missing_med$incident_location == "Europe"), "Europe Route","Other")))))))


```

Now map the migrant deaths by the new route feature 
```{r}

color <- colorFactor(topo.colors(8), missing_med$route)


leaflet() %>% 
  addTiles() %>%
  addCircles(data = missing_med, group = "route", lng = ~lon, lat = ~lat, weight = 4, color = ~color(route), stroke = TRUE,
    radius = ~dead * 500, popup = ~route) 
```
Let's narrow our focus a bit to look at incidents of drowning and those that are Syrian.  

Here we create a few more features for our data set.
```{r}
# create a new data frame just in case we want to go back and try something else.  
df <- missing_med 

# create a new True/False Feature for drowning 
df$drown <- ifelse((missing_med$cause_death == "Drowning"), "TRUE", "FASLE")

# The nationality column is pretty messey as well. We'll try a simple fix, but given how messy this 
# column was earlier we'd urge caution in a real world analysis setting. 
df$nationality[startsWith(df$nationality, "Syr")] <- "Syrian"

# now view the missing and dead Syrians in a table 
nation <- df %>% 
    group_by(nationality == 'Syrian') %>% 
    summarise(sum(missing), sum(dead))

nation


```


```{r}
# create a true/false feature for Syrian 
df$syrian <- ifelse((df$nationality == "Syrian"), "TRUE", "FASLE")

# check the new syrian column to make sure it looks like what we did above 
nation <- df %>% 
    group_by(syrian == 'TRUE') %>% 
    summarise(sum(missing), sum(dead))

nation
```
### Syrian Migrant Deaths by Route 
```{r}
missing_syrians <- subset(df, syrian == 'TRUE')


color <- colorFactor(topo.colors(8), missing_syrians$route)


leaflet() %>% 
  addTiles() %>%
  addCircles(data = missing_syrians, group = "route", lng = ~lon, lat = ~lat, weight = 4, color = ~color(route), stroke = TRUE,
    radius = ~dead * 500, popup = ~route)
 

```

